---
title: "Validation & Limitations"
description: "Validation layer documentation, model governance, and study limitations"
---

## Validation Layer

All scripts in this project pass through a shared validation layer defined in
[`R/validation.R`](R/validation.R). The layer enforces the following checks
before any analysis runs:

| Check | Function | What It Guards |
|---|---|---|
| Column presence | `validate_columns()` | Ensures required variables exist |
| Variable types | `validate_types()` | Ensures factors/numerics are correctly typed |
| Two-group structure | `validate_two_groups()` | Ensures exactly 2 levels for comparisons |
| Sample size | `validate_sample_size()` | Minimum 30 overall, 10 per group |
| Missingness | `validate_missingness()` | Warns if any column exceeds 5% NAs |
| Range bounds | `validate_ranges()` | Grades 0–20, age 15–22, absences 0–100 |
| Causal language | `validate_causal_claims()` | Warns on deterministic causal wording |

---

## Model Governance

### Baseline Comparison

```{r}
#| label: baseline
metrics <- read.csv("models/model_metrics.csv")
acc  <- metrics$value[metrics$metric == "Accuracy"]
n    <- nrow(read.csv("data/processed/student_clean.csv"))
at_risk_pct <- mean(read.csv("data/processed/student_clean.csv")$at_risk == "At Risk")
majority_class_acc <- round(max(at_risk_pct, 1 - at_risk_pct), 4)

cat(sprintf(
  "Logistic regression accuracy: %.4f\nMajority class baseline:       %.4f\nImprovement over baseline:     %.4f",
  acc, majority_class_acc, acc - majority_class_acc
))
```

### Fairness Checks

```{r}
#| label: fairness
library(dplyr, warn.conflicts = FALSE)
df    <- read.csv("data/processed/student_clean.csv", stringsAsFactors = TRUE)
model <- readRDS("models/student_risk_model.rds")

df$pred_prob  <- predict(model, newdata = df, type = "response")
df$pred_class <- ifelse(df$pred_prob > 0.5, "At Risk", "Low Risk")

# Accuracy by sex
acc_by_sex <- df |>
  group_by(sex) |>
  summarise(
    n          = n(),
    accuracy   = round(mean(pred_class == as.character(at_risk)), 4),
    pct_at_risk_actual    = round(mean(at_risk == "At Risk") * 100, 1),
    pct_at_risk_predicted = round(mean(pred_class == "At Risk") * 100, 1),
    .groups = "drop"
  )

knitr::kable(acc_by_sex,
             caption = "Model accuracy and at-risk rates by sex",
             digits = 4)

# Accuracy by school
acc_by_school <- df |>
  group_by(school) |>
  summarise(
    n          = n(),
    accuracy   = round(mean(pred_class == as.character(at_risk)), 4),
    pct_at_risk_actual    = round(mean(at_risk == "At Risk") * 100, 1),
    pct_at_risk_predicted = round(mean(pred_class == "At Risk") * 100, 1),
    .groups = "drop"
  )

knitr::kable(acc_by_school,
             caption = "Model accuracy and at-risk rates by school",
             digits = 4)
```

---

## Study Limitations

1. **Observational design**: The dataset was collected retrospectively. No
   randomised experiment was conducted. All associations may be confounded
   by unmeasured variables (e.g., teacher quality, individual motivation).

2. **Single subject, single cohort**: Results apply to Mathematics students
   in two Portuguese schools in 2005–2006. Generalisability to other subjects,
   countries, or time periods is unknown.

3. **Small sample size (N = 395)**: The dataset is too small for robust
   subgroup analysis or complex modelling. Confidence intervals are wide.

4. **Self-reported predictors**: Variables such as `studytime`, `freetime`,
   `goout`, and alcohol consumption are self-reported and subject to response bias.

5. **Binary risk threshold**: The cut-off G3 < 10 is a policy threshold, not
   a natural category. Students scoring 10 are treated identically to those
   scoring 20, despite meaningful differences.

6. **Model exclusions**: G1 and G2 are excluded to avoid data leakage.
   A model that includes them would have higher accuracy but limited utility
   for early intervention.

7. **No temporal validation**: The model is evaluated on a held-out test set
   from the same cohort, not on future cohorts (prospective validation).

---

## Responsible Use Statement

This project is designed as a **decision-support tool**, not an automated
decision-maker. It should be used to:

- ✅ Help counsellors prioritise their time
- ✅ Surface students who may otherwise go unnoticed
- ✅ Stimulate conversations about intervention options

It should **not** be used to:

- ❌ Automatically enrol students in programmes without consent
- ❌ Share predictions with students or parents without professional review
- ❌ Replace qualitative judgement with algorithmic scoring
- ❌ Infer causation from association-based predictions
